Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-18 manuscript.tex
Updates from Overleaf

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}. A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}. Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}. Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}. The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides \citep{Sanstead2016}. Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

@@

The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent. \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries. Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library \citep{McGibbon2015MDTraj:Trajectories}. Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. Additional details on the VAMP-2 score can be found in the supplemental information. By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure. Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.

@@

% need some more detail here -- talk to Andy about this
MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.

\subsection{\label{sec:methods}Diffusion Maps}
Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}. They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time. We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies \citep{Shmilovich, Wang2017MesoscaleLandscapes}. We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.

@@

As an initial validation of 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously \citep{Sanstead2018DirectDehybridization}. We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation \citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC}. We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (figure \ref{fig:relaxation-comparison})). At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}. All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}. Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends\citep{Santalucia2004TM}. The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models \citep{Michele2014EHybridization}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

>>>>

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications.\citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine} In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation.\citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment} A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects.\citep{SantaLucia1998AThermodynamics, Santalucia2004TM} Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications.\citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology} Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood.\citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon} Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition.\citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium.\citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization} More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers.\citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution} Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail.\citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques.\citep{Phys2014} Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling.\citep{Hinckley2013AnHybridization, Schmitt2013ExploringSurface, Sambriski2009, Sambriski2009SequencePathways, Hoefert2011MolecularOligonucleotides,Romano2013DNADependence, Hinckley2014Coarse-grainedEffects} Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative.\citep{Araque2016LatticeCooperativity, Phys2019} Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events.\citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding} Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics.\citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence} Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies.\citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019} The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies.\citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix} Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides.\citep{Sanstead2016} Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above.\citep{Sanstead2016} We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature.\citep{Hinckley2013AnHybridization} We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics.\citep{Pande2010EverythingAsk} MSMs have recently been implemented to study mechanisms of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies.\citep{Jin2019, Xiao2019} \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA oligonucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

@@

The 3spn2 coarse grained model was designed to accelerate DNA MD relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths.\citep{Hinckley2013AnHybridization} Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients.\citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects} The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm.\citep{Sanstead2016} We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent.\citep{Schneider1978Molecular-dynamicsTransitions} Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries. Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library.\citep{McGibbon2015MDTraj:Trajectories} Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction.\citep{Sengupta2019AutomatedSelf-assembly} The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters.\citep{WuVariationalData, Mardt2018VAMPnetsKinetics} Additional details on the VAMP-2 score can be found in the supplemental information. By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure. Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVs as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator.\citep{Chen} The method is a descendant of VAMPnets\citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL).\citep{Li2017ExtendedOperator} The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated.\citep{Noe2013ASystems} The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs.\citep{Sidky, Pande2010EverythingAsk} SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein.\citep{Chen, Sidky}

@@

% need some more detail here -- talk to Andy about this
MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way.\citep{Phys2011MarkovValidation, Husic2018MarkovScience} Kinetically similar conformations are first discretized into microstates, and the conditional probability of transitions between microstates is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible.\citep{Pande2010EverythingAsk} Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they function as an optimal MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence.\citep{Scherer2015PyEMMAModels} In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments.\citep{Sidky} The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and is discussed more in depth in the results and supplemental information.

\subsection{\label{sec:methods}Diffusion Maps}
Diffusion maps generate a low-dimensional non-linear embedding of data into high variance structural modes.\citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps} They are well-suited to find subtle differences in molecular conformations that are temporally disconnected or that interchange faster than the MSM lag time. We used reciprocal inter-molecular base distances as inputs to diffusion maps, and tuned $\epsilon$ and $\alpha$ parameters in accordance with previous studies.\citep{Shmilovich, Wang2017MesoscaleLandscapes} We found that the choice of these parameters mainly served to condense or stretch the low dimensional space and had a minimal qualitative impact on salient results.

@@

As an initial validation of 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) IR experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously.\citep{Sanstead2018DirectDehybridization} We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We calculated relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 1.3 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation.\citep{Craig1971ElaxationOligon,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. We also noted an acceleration of about one order of magnitude compared to experiment, although this factor is sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw strong agreement between the experimental data and simulated relaxation fits (figure \ref{fig:relaxation-comparison})). At lower temperatures, T-jump slow responses likely contain a mixture of dissociation and hybridization dynamics, therefore rates do not drop off to the same extent as our analysis, which considers dissociation alone. GC-core showed the largest deviation from experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

@@

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds.\citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC} All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations.\citep{PinamontiTheModels, Zgarbova2014BaseRNA} Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until duplex terminal ends splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.

@@

Nearest neighbor (NN) models have been experimentally optimized to account for stacking contributions of native pairs, internal mismatches, and dangling ends.\citep{Santalucia2004TM} The effects of inert tails -- free bases that extend beyond the dangling end -- have been shown to destabilize the duplex but are not included in NN models.\citep{Michele2014EHybridization} It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. In particular, we examine  the out-of-register AT-all and GC-end states which are amenable to NN comparisons (figure \ref{fig:NN_table}). For AT-all, NN calculations  predict that conformations in the 5' shifted states (5S2 and 5S4) are more energetically favorable than those in the 3' shifted state (3S2 and 3S4). We find qualitative agreement to our macrostate free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to our simulation results, however, the NN model predicts the 3S2 state to be more stable than 5S2 state. Moreover, we do not observe a stable 3S4 cluster for GC-end. This indicates that 5' vs. 3' inert tail differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone.\citep{Doktycz1990ThermodynamicATGC}
